Minimizing Time Required For Graduation

### 2010-12-04 paper.tex
Now spell checked

\section{Introduction} \subsection{Background} Registration can be a stressful
time for every student at the Univeristy of Washington. Without deliberate
planning, students must decide which classes to take on a quarterly basis. The
decision process involves considering many factors, such as quarterly class
offerings, the amount of credits, the prerequisites of the classes, time
conflicts, and how the choices in current quarter affect future quarters. The
University of Washington registration system provides a schedule finder;
however, it only takes into account time conflicts, and does not give advice on
how to pick the most ideal schedule. The situation described above makes it
difficult for students advising faculties. A student's graduation time may be
extended because of his or her inability to take a class that has limited
offerings during that particular year, for reasons such as not satisfying the
prerequisites of the class. The advising office’s workload can also be increased
due to the changing plans and majors of students and their need for repeated
advising.

\subsection{Goal} Our goal is to alleviate the problematic situation by creating
a mathematical algorithm that accomplishes the following goal: given a student's
current state and preferred credit hours per quarter, produce a class schedule
that minimizes the number of quarters until graduation, eliminates time
conflicts, and eliminates prerequiste conflicts.

\subsection{Importance} The implementation of the algorithm as part of the
school’s registration system will benefit the students by enabling them to map
out their quarterly class schedules until graduation, and to preview their ideal
schedule, the length of time until graduation, and the credit hours per quarter.
This will lead to better decision making for what majors to pursue, and given
the student's current state, which schedules are the most ideal in terms of
saving time and resources. This will also help studentssave money, as they will
be in school for fewer quarters. The university also benefits from the reduced
labor cost and increased efficiency of the advising offices.

@@

We must now also define some variables used throughout our paper.
\begin{itemize} \item Let $S$ be the set of courses required to be taken in
order to graduate. $S$ can be thought of as our goal state, and can also contain
elective courses a student simply wishes to take.  \item Let $T$ be the set of
courses a given student has already completed. If no courses have been taken,
this set may be empty.  \item Let $U$ be the set of all courses a student could
possibly ever take.  \item Let $P$ be a set of rules specifying course
prerequistes.  \item Let $l$ be the maximum number of credits a student wishes
to take during any given quarter.  \item Let $O$ be the set of all 50-minute
time blocks between 8:30am and 7:30pm from Monday through Friday.  \end{itemize}

\subsection{Representing Prerequisites} Initially, we hoped to represent course
prerequistes using a directed acyclic graph, in which each node was a course,
and edges between nodes $i$ and $j$ if and only if $i$ was a prerequisite for
$j$.  This method was used by Pritsker et. al. and Xiangtong et. al. However, it
does not work in our model, and courses can have a complicated set of multiple
prerequisites. For example, {\it MATH327} requires both $MATH126$ and $MATH300$,
or just $MATH136$. Let this graph be called $G_{prereq}$. Due to these
complications, a $G_{prereq}$ alone will not suffice.

To solve this issues, we introduce a set of rules for each course. Each rule is
a boolean expression represented in propositional logic in disjunctive normal
form. For example, the rule assoicated with $MATH327$ is $$ \text{{\it MATH327:
(MATH126 AND MATH300) OR MATH136}}.  $$ As previously stated, we denote the set
of all rules for all classes as $P$.

@@

\subsection{Initial Model and Tracability} With a static schedule it would be
possible to deterministically plan an entire schedule from start to finish,
unfortunately we have neither the computational resources nor the luxury of
a static schedule. From the student’s point of view, although there are some
classes that are offered every quarter, senior electives are not, and sometimes
it cannot be predicted more than one quarter in advance when a senior elective
will be offered. Even given perfect information, where we know the possible
class schedules for every quarter we face another problem; finding the best
possible schedule reduces to enumerating possible class schedules for every
quarter until the graduation requirements are satisfied. Given that we cannot
predict in advance the conflicts between classes, we cannot even determine the
branching factor of the enumeration tree nor its depth. Since we can only
determine one quarter in advance, this suggests approaching the problem by using
heuristics to pick the best possible classes each quarter.

@@

\section{Rolling Horizons Approach} We adapt a rolling horizons approach which
uses a heuristic to pick the best possible schedule for any given quarter, given
the classes offered during the quarter and the courses the student has
previously taken. We use a greedy algorithm that functions on a per-quarter
basis to accomplish this. The greedy algorithm greatly improves the tracability
of course scheduling over an initially unknown number of quarters.

@@

\subsubsection{Procedue} We present the procedure for popluating the set $L$ as
follows: \begin{enumerate} \item For each rule in $P$ we iterate over each
proposition in the rule, $p$, and check if $p \in T$. If $p \in T$ then we say
$p$ evaluates to true; otherwise, $p$ evaluates to false. We recall that each
rule is associated with a given course. If the entire rule evaluates to true,
then we add the course associated with it to $L$. Once this is complete, $L$
will consist of every course that could be taken based solely on prerequisites
that we have fulfilled.  \item We now prune this set by removing all courses
that are not offered in the current quarter. $L$ will necessarily now only
contain classes that we can take given the current quarter and the classes we
have already taken.  \item We now construct a confict graph, $G_{conflict}$,
where every course offering in $L$ is a node, and there is an edge between two
nodes $i$ and $j$ if and only if the time of course offering $i$ overlaps the
time of course offering $j$.  \end{enumerate}

\subsubsection{Analysis} The selection procedure involves a lot of set queries
and iteration, but in asymptotic running time the algorithm is relatively quick.
Let $n$ be the total number of classes ever available to take, or in other
words, $n = \|U\|$. In the worst case, $P$ has a complex rule for all $n$
courses, and checking each rule in $P$ is linear in the number of propositions.
However, we know {\it a priori} from analysis of our dataset that the largest
rule has twelve propositions, so we can bound the rule check time by a constant,
$C$, and therefore step one can be done in $O(Cn)$ time.

@@

Since we are dealing with asymptotic running time we will drop the constant
terms and combining all resulsts yields an overall runtime of $O(n)$.

\subsection{Selection of Quarterly Classes} \subsubsection{Selection Prodcedure}
At this point in time, we know which course offerings are available for the
given student to register for, and we must now pick the most optimal ones. There
a several ways we could view the problem of determining which classes to take
each quarter. We will initially encode the class selection as a binary integer
programming problem in spirit of the methods presented by Dinkel et al
\cite{dinel:scheduling}. and Pritsker et al \cite{prisker:informs}. At this
point we have the set $L$ containing possible classes we that we need to take.

\subsubsection{Heuristic Value Function} In order to determine which classes
should be picked in a given schedule, we use a heuristic function which assigns
each course offering a value. The higher the value of this function for a given
course offering is, the more likely the given course offering will be placed in
the schedule. Recalling that our goal is to minimize the total number of
quarters taken (and implicitly, maximize the total number of courses taken per
quarter), our heuristic function should clearly give higher value to course
offerings that will help us achieve this goal. One component to the heuristic
function then must be the number of dependent courses a given course offering
has. If a course satisfies many prerequisties, then we know that there are many
courses that we cannot complete without it, so we want our heuristic function to
favor it.

Additionally, we know that not every course is offered every quarter. If
a course is not offered until the following year, we could potentially add an
entire year to the schedule. This is a situation we hope to avoid. Our heuristic
function must therefore consider how often and how far in the future a given
course is offered. For example, if we are deciding between two course offering
that are identical except one is ofference once a year, and the other is offered
three times a year, we want to take the course only offered once a year, as we
won't have the oppurtunity to take it for another entire year.

We these considerations in mind, we begin to develop our heuristic function. In
order to define our heuristic function, we must first define some auxilary
functions on course offering.  \begin{itemize} \item Let $TimeFromNow(c, q)$ be
difference in quarters between the current quarter, $q$, and the next quarter
a given course offering, $c$, is offered. As there are four quarters, the range
of this function is $\{0, 1, 2, 3\}$.  \item Let $Quarters(c)$ be the number of
quarters per year a given course, $c$, is offered. As there are four quarters,
the range of this function is $\{0, 1, 2, 3, 4\}$.  \item Let $Dependents(c)$ be
the number of course that either directly or indirectly require a given course,
$c$. In our final draft, we will present a diagram illustrating this value more
clearly.  \end{itemize}

@@

\subsubsection{Formulation as Binary Integer Program} We can now formulate our
problem as a binary integer linear programming probelm. Before stating the
linear program, we must first define some variables used in it: \\
\begin{itemize} \item $ X_i = \left\{ \begin{array}{lr} 1 \text{ if we take
} X_i\\ 0 \text{ if we don't take } X_i \end{array} \right. $ \item For all $t
\in O$, we let $t = 1$.  \item Let $m = \|L\|$.  \end{itemize} We now state our
problem as follows:\\ Maximize: $$ \text{Value of Quarter} = \sum_{i=1}^m
Value(c_i) * X_i $$ Subject To: $$ \sum_{i=1}^m Weight(c_i) * X_i \leq l, $$ $$
\forall t_j \in O, \sum_{i=1}^m t_j * X_i \leq 1, $$ $$ X_i \in \{0, 1\} $$ The
linear program can be interpreted as follows: \begin{itemize} \item The value of
a quarter is the total number of courses taken that quarter.  \item In each
quarter, the total number of credits cannot exceed the specified limit.  \item
Only one course can be taken in one particular time slot \item A course can only
be taken or not taken. In other words, decision variables are binary.  \item
Each non-zero descision variable in the optimal solution represents a class we
will take.  \end{itemize}

@@

\subsubsection{Relaxation and Implicit Enumeration} \begin{enumerate} \item
Develop upper bound on value by solving a continuous version knapsack problem
generated by a Lagrangean relaxation of our problem. Let $X_i$ be a continuous
variable with $0 \leq X_i \leq 1$, and shift the time constraints into the
objective function by using a penalty of $\lambda$ when the constraint is
violated.  \item Use the upper bound developed above with the pruning rules
suggested by Yamada et al \cite{yamada:heuristic} to do implicit enumeration of
the possible choices.  \end{enumerate}

>>>>

\section{Introduction} \subsection{Background} Registration can be a stressful
time for every student at the University of Washington. Without deliberate
planning, students must decide which classes to take on a quarterly basis. The
decision process involves considering many factors, such as quarterly class
offerings, the amount of credits, the prerequisites of the classes, time
conflicts, and how the choices in current quarter affect future quarters. The
University of Washington registration system provides a schedule finder;
however, it only takes into account time conflicts, and does not give advice on
how to pick the most ideal schedule. The situation described above makes it
difficult for students advising faculties. A student's graduation time may be
extended because of his or her inability to take a class that has limited
offerings during that particular year, for reasons such as not satisfying the
prerequisites of the class. The advising office’s workload can also be increased
due to the changing plans and majors of students and their need for repeated
advising.

\subsection{Goal} Our goal is to alleviate the problematic situation by creating
a mathematical algorithm that accomplishes the following goal: given a student's
current state and preferred credit hours per quarter, produce a class schedule
that minimizes the number of quarters until graduation, eliminates time
conflicts, and eliminates prerequisite conflicts.

\subsection{Importance} The implementation of the algorithm as part of the
school’s registration system will benefit the students by enabling them to map
out their quarterly class schedules until graduation, and to preview their ideal
schedule, the length of time until graduation, and the credit hours per quarter.
This will lead to better decision making for what majors to pursue, and given
the student's current state, which schedules are the most ideal in terms of
saving time and resources. This will also help students save money, as they will
be in school for fewer quarters. The university also benefits from the reduced
labor cost and increased efficiency of the advising offices.

@@

We must now also define some variables used throughout our paper.
\begin{itemize} \item Let $S$ be the set of courses required to be taken in
order to graduate. $S$ can be thought of as our goal state, and can also contain
elective courses a student simply wishes to take.  \item Let $T$ be the set of
courses a given student has already completed. If no courses have been taken,
this set may be empty.  \item Let $U$ be the set of all courses a student could
possibly ever take.  \item Let $P$ be a set of rules specifying course
prerequisites.  \item Let $l$ be the maximum number of credits a student wishes
to take during any given quarter.  \item Let $O$ be the set of all 50-minute
time blocks between 8:30am and 7:30pm from Monday through Friday.  \end{itemize}

\subsection{Representing Prerequisites} Initially, we hoped to represent course
prerequisites using a directed acyclic graph, in which each node was a course,
and edges between nodes $i$ and $j$ if and only if $i$ was a prerequisite for
$j$.  This method was used by Pritsker et. al. and Xiangtong et. al. However, it
does not work in our model, and courses can have a complicated set of multiple
prerequisites. For example, {\it MATH327} requires both $MATH126$ and $MATH300$,
or just $MATH136$. Let this graph be called $G_{prereq}$. Due to these
complications, a $G_{prereq}$ alone will not suffice.

To solve this issues, we introduce a set of rules for each course. Each rule is
a boolean expression represented in propositional logic in disjunctive normal
form. For example, the rule associated with $MATH327$ is $$ \text{{\it MATH327:
(MATH126 AND MATH300) OR MATH136}}.  $$ As previously stated, we denote the set
of all rules for all classes as $P$.

@@

\subsection{Initial Model and Tractability} With a static schedule it would be
possible to deterministically plan an entire schedule from start to finish,
unfortunately we have neither the computational resources nor the luxury of
a static schedule. From the student’s point of view, although there are some
classes that are offered every quarter, senior electives are not, and sometimes
it cannot be predicted more than one quarter in advance when a senior elective
will be offered. Even given perfect information, where we know the possible
class schedules for every quarter we face another problem; finding the best
possible schedule reduces to enumerating possible class schedules for every
quarter until the graduation requirements are satisfied. Given that we cannot
predict in advance the conflicts between classes, we cannot even determine the
branching factor of the enumeration tree nor its depth. Since we can only
determine one quarter in advance, this suggests approaching the problem by using
heuristics to pick the best possible classes each quarter.

@@

\section{Rolling Horizons Approach} We adapt a rolling horizons approach which
uses a heuristic to pick the best possible schedule for any given quarter, given
the classes offered during the quarter and the courses the student has
previously taken. We use a greedy algorithm that functions on a per-quarter
basis to accomplish this. The greedy algorithm greatly improves the tractability
of course scheduling over an initially unknown number of quarters.

@@

\subsubsection{Procedure} We present the procedure for populating the set $L$ as
follows: \begin{enumerate} \item For each rule in $P$ we iterate over each
proposition in the rule, $p$, and check if $p \in T$. If $p \in T$ then we say
$p$ evaluates to true; otherwise, $p$ evaluates to false. We recall that each
rule is associated with a given course. If the entire rule evaluates to true,
then we add the course associated with it to $L$. Once this is complete, $L$
will consist of every course that could be taken based solely on prerequisites
that we have fulfilled.  \item We now prune this set by removing all courses
that are not offered in the current quarter. $L$ will necessarily now only
contain classes that we can take given the current quarter and the classes we
have already taken.  \item We now construct a conflict graph, $G_{conflict}$,
where every course offering in $L$ is a node, and there is an edge between two
nodes $i$ and $j$ if and only if the time of course offering $i$ overlaps the
time of course offering $j$.  \end{enumerate}

\subsubsection{Analysis} The selection procedure involves a lot of set queries
and iteration, but in asymptotic running time the algorithm is relatively quick.
Let $n$ be the total number of classes ever available to take, or in other
words, $n = \|U\|$. In the worst case, $P$ has a complex rule for all $n$
courses, and checking each rule in $P$ is linear in the number of propositions.
However, we know {\it a priori} from analysis of our data set that the largest
rule has twelve propositions, so we can bound the rule check time by a constant,
$C$, and therefore step one can be done in $O(Cn)$ time.

@@

Since we are dealing with asymptotic running time we will drop the constant
terms and combining all results yields an overall run time of $O(n)$.

\subsection{Selection of Quarterly Classes} \subsubsection{Selection Procedure}
At this point in time, we know which course offerings are available for the
given student to register for, and we must now pick the most optimal ones. There
a several ways we could view the problem of determining which classes to take
each quarter. We will initially encode the class selection as a binary integer
programming problem in spirit of the methods presented by Dinkel et al
\cite{dinel:scheduling}. and Pritsker et al \cite{prisker:informs}. At this
point we have the set $L$ containing possible classes we that we need to take.

\subsubsection{Heuristic Value Function} In order to determine which classes
should be picked in a given schedule, we use a heuristic function which assigns
each course offering a value. The higher the value of this function for a given
course offering is, the more likely the given course offering will be placed in
the schedule. Recalling that our goal is to minimize the total number of
quarters taken (and implicitly, maximize the total number of courses taken per
quarter), our heuristic function should clearly give higher value to course
offerings that will help us achieve this goal. One component to the heuristic
function then must be the number of dependent courses a given course offering
has. If a course satisfies many prerequisites, then we know that there are many
courses that we cannot complete without it, so we want our heuristic function to
favor it.

Additionally, we know that not every course is offered every quarter. If
a course is not offered until the following year, we could potentially add an
entire year to the schedule. This is a situation we hope to avoid. Our heuristic
function must therefore consider how often and how far in the future a given
course is offered. For example, if we are deciding between two course offering
that are identical except one is offered once a year, and the other is offered
three times a year, we want to take the course only offered once a year, as we
won't have the opportunity to take it for another entire year.

We these considerations in mind, we begin to develop our heuristic function. In
order to define our heuristic function, we must first define some auxiliary
functions on course offering.  \begin{itemize} \item Let $TimeFromNow(c, q)$ be
difference in quarters between the current quarter, $q$, and the next quarter
a given course offering, $c$, is offered. As there are four quarters, the range
of this function is $\{0, 1, 2, 3\}$.  \item Let $Quarters(c)$ be the number of
quarters per year a given course, $c$, is offered. As there are four quarters,
the range of this function is $\{0, 1, 2, 3, 4\}$.  \item Let $Dependents(c)$ be
the number of course that either directly or indirectly require a given course,
$c$. In our final draft, we will present a diagram illustrating this value more
clearly.  \end{itemize}

@@

\subsubsection{Formulation as Binary Integer Program} We can now formulate our
problem as a binary integer linear programming problem. Before stating the
linear program, we must first define some variables used in it: \\
\begin{itemize} \item $ X_i = \left\{ \begin{array}{lr} 1 \text{ if we take
} X_i\\ 0 \text{ if we don't take } X_i \end{array} \right. $ \item For all $t
\in O$, we let $t = 1$.  \item Let $m = \|L\|$.  \end{itemize} We now state our
problem as follows:\\ Maximize: $$ \text{Value of Quarter} = \sum_{i=1}^m
Value(c_i) * X_i $$ Subject To: $$ \sum_{i=1}^m Weight(c_i) * X_i \leq l, $$ $$
\forall t_j \in O, \sum_{i=1}^m t_j * X_i \leq 1, $$ $$ X_i \in \{0, 1\} $$ The
linear program can be interpreted as follows: \begin{itemize} \item The value of
a quarter is the total number of courses taken that quarter.  \item In each
quarter, the total number of credits cannot exceed the specified limit.  \item
Only one course can be taken in one particular time slot \item A course can only
be taken or not taken. In other words, decision variables are binary.  \item
Each non-zero decision variable in the optimal solution represents a class we
will take.  \end{itemize}

@@

\subsubsection{Relaxation and Implicit Enumeration} \begin{enumerate} \item
Develop upper bound on value by solving a continuous version knapsack problem
generated by a Lagrangian relaxation of our problem. Let $X_i$ be a continuous
variable with $0 \leq X_i \leq 1$, and shift the time constraints into the
objective function by using a penalty of $\lambda$ when the constraint is
violated.  \item Use the upper bound developed above with the pruning rules
suggested by Yamada et al \cite{yamada:heuristic} to do implicit enumeration of
the possible choices.  \end{enumerate}
